---
title: "Calling R in Haskell"
date: "2016-09-03"
output: html_document
---

<link rel="stylesheet" type="text/css" href="assets/css/hscolour.css">

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "./assets/haskell/")
## a small modification of the haskell engine provided by knitr:
##   (to get rid of the option ':set +m')
knitr::knit_engines$set(ghc = function (options) 
                  {
                    engine = options$engine
                    f = basename(tempfile(engine, ".", ".txt"))
                    writeLines(options$code, f)
                    on.exit(unlink(f))
                    code = paste(f, options$engine.opts)
                    cmd = options$engine.path 
                    out = if (options$eval) {
                      message("running: ", cmd, " ", code)
                      tryCatch(system2(cmd, code, stdout = TRUE, stderr = FALSE,
                                       env = options$engine.env), error = function(e) {
                                         if (!options$error)
                                           stop(e)
                                         paste("Error in running command", cmd)
                                       })
                    }
                    else ""
                    if (!options$error && !is.null(attr(out, "status"))) 
                      stop(paste(out, collapse = "\n"))
                    knitr::engine_output(options, options$code, out)
                  }
)
## chunk options
knitr::opts_chunk$set(engine = 'ghc', 
                      engine.path = 'ghcscriptrender', 
                      engine.opts = '-s --fragment -p /home/stla/.cabal-sandbox/x86_64-linux-ghc-7.10.3-packages.conf.d',
                      echo = FALSE,
                      results = 'asis', 
                      cache = TRUE)
```


This is a demo of  the [inline-R](https://tweag.github.io/HaskellR/) Haskell library, allowing to call R from Haskell. 

In order to call R in GHCi, you have to run this preamble before:

```{r engine.opts="--module --fragment"}
:set -XDataKinds
:set -XGADTs
:set -XPartialTypeSignatures
:set -XOverloadedLists
:set -XQuasiQuotes
:set -XScopedTypeVariables
:set -XTemplateHaskell
:set -XViewPatterns
import Data.Int (Int32)
import H.Prelude.Interactive as H.Prelude
import Language.R.Instance as R
R.initialize R.defaultConfig
```

Therefore, since I use [knitr+ghcscriptrender](http://stla.github.io/stlapblog/posts/ghcscriptrender.html) to write this article, I save the following code in a file `Rpreamble.hs`:

```{r engine.opts="--module --fragment"}
import Data.Int (Int32)
import H.Prelude.Interactive as H.Prelude
import Language.R.Instance as R

main :: IO()
main = do
  R.initialize R.defaultConfig
  return()
```

And now I can call R like this:

```{r}
:load Rpreamble.hs
:set -XTemplateHaskell -XQuasiQuotes -XScopedTypeVariables -XDataKinds -XGADTs -XPartialTypeSignatures -XOverloadedLists -XViewPatterns
-- perform 1+1 in R:
x <- [r| 1 + 1 |]
fromSomeSEXP x :: Double
-- integers from 1 to 3:
x <- [r| 1:3 |]
fromSomeSEXP x :: [Int32]
```

One can run several commands in `[r| ...|]`:

```{r}
:load Rpreamble.hs
:set -XTemplateHaskell -XQuasiQuotes -XScopedTypeVariables -XDataKinds -XGADTs -XPartialTypeSignatures -XOverloadedLists -XViewPatterns
-- separate with a semi-colon:
x <- [r| a <- 2; 3+a |]  
fromSomeSEXP x :: Double
-- or multiline:
:{
x <- [r| a <- 2
3+a |]
:}
fromSomeSEXP x :: Double
```

One can load and use a package:

```{r}
:load Rpreamble.hs
:set -XTemplateHaskell -XQuasiQuotes -XScopedTypeVariables -XDataKinds -XGADTs -XPartialTypeSignatures -XOverloadedLists -XViewPatterns
-- load and use the jsonlite package
:{
x <- [r| library(jsonlite)
dat <- data.frame(name=c('Joe', 'Bill'), age=c(23,40))
toJSON(dat) |]
:}
fromSomeSEXP x :: String
-- But I prefer this way sometimes:
:{
x <- [r| dat <- data.frame(name=c('Joe', 'Bill'), age=c(23,40))
jsonlite::toJSON(dat) |]
:}
fromSomeSEXP x :: String
```

The most interesting feature is the ability to pass a Haskell variable in R:

```{r}
:load Rpreamble.hs
:set -XTemplateHaskell -XQuasiQuotes -XScopedTypeVariables -XDataKinds -XGADTs -XPartialTypeSignatures -XOverloadedLists -XViewPatterns
-- 
let a = 2::Double
x <- [r| 3 + a_hs|]
fromSomeSEXP x :: Double
```

The `p` function displays the output of R's `print` function:

```{r}
:load Rpreamble.hs
:set -XTemplateHaskell -XQuasiQuotes -XScopedTypeVariables -XDataKinds -XGADTs -XPartialTypeSignatures -XOverloadedLists -XViewPatterns
-- print
p [r| 1+1 |]
```

The `jsonlite` package allows to get arrays and lists from Haskell:

```{r show, engine.opts = '--fragment -p /home/stla/.cabal-sandbox/x86_64-linux-ghc-7.10.3-packages.conf.d'}
:load Rpreamble.hs
:set -XTemplateHaskell -XQuasiQuotes -XScopedTypeVariables -XDataKinds -XGADTs -XPartialTypeSignatures -XOverloadedLists -XViewPatterns
-- an array:
let a = [[1,2],[3,4]]::[[Int]]
let b = show a
p [r| jsonlite::fromJSON(b_hs) |]
-- a list:
let a = [[1,2],[3,4,5]]::[[Int]]
let b = show a
p [r| jsonlite::fromJSON(b_hs) |]
```

And dataframes, with the help of the [aeson library](https://artyom.me/aeson).

```{r dataframe, engine.opts = '--fragment -p /home/stla/.cabal-sandbox/x86_64-linux-ghc-7.10.3-packages.conf.d'}
:load Rpreamble.hs
:set -XTemplateHaskell -XQuasiQuotes -XScopedTypeVariables -XDataKinds -XGADTs -XPartialTypeSignatures -XOverloadedLists -XViewPatterns
-- get the JSON representation of a dataframe:
import Data.Aeson
import GHC.Generics
:set -XOverloadedStrings -XDeriveGeneric
data Person = Person { name :: String, age :: Int } deriving (Show, Generic)
instance ToJSON Person
let persons = [Person "Joe" 21, Person "Bill" 40]::[Person]
encode persons
-- pass it to R:
import qualified Data.Text.Lazy.Encoding as T
import qualified Data.Text.Lazy as T
let b = T.unpack $ T.decodeUtf8 $ encode persons
p [r| jsonlite::fromJSON(b_hs) |]
```

