---
title: "The expansions package: integer and float expansions, and odometers"
date: "2016-08-13"
output: html_document
---

```{r setup, include=FALSE}
library(expansions)
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, fig.path="./assets/fig/expansions-")
```


I released the `expansions` package for R [on github](https://github.com/stla/expansions). 
It works on Linux only. 

## Expansions of integer and real numbers

### Conversion of decimal integers to integer base

This is the role of the `intAtBase` function:

```{r}
intAtBase(14, base=3)
2*1 + 1*3 + 1*3^2
```

### Cantor expansions, or "ary" expansions 

The `intToAry` function performs the Cantor expansion of an integer. I explained the Cantor expansion in [a previous article](http://stla.github.io/stlapblog/posts/Greedy.html). 

```{r}
# (3,4,7)-expansion of 77:
intToAry(77, c(3,4,7))
2*1 + 1*3 + 6*(3*4)
```

```{r}
# Cartesian product {0,1}x{0,1}:
sapply(0:3, function(x) intToAry(x, sizes=c(2,2)))
```

It is implemented with `Rcpp`, and it is fast. 

### Float expansions

The `floatExpand01` function returns the expansion of a real number between $0$ and $1$ to its expansion in a given integer base. For example:

```{r}
floatExpand01(0.625, base=2)
```

It means that $0.625 = 1\times \frac{1}{2} + 0 \times \frac{1}{2^2} + 1 \times \frac{1}{2^2}$.

The `floatExpand` function returns a list representing the expansion of a positive number in scientific notation:
$$
x = 0.d_1d_2\ldots d_n \times \text{base}^e.
$$
the digits $d_1$, $\ldots$, $d_n$ are given in the first component of the list and the exponent $e$ is given in the second one.
For example:


```{r}
floatExpand(1.125, base=2)
(1*1/2 + 0*1/2^2 + 1*1/2^3) * 2^1
```


## Odometers and addition of adic integers

The odometer is the action $x \mapsto x+1$ on the group of $b$-adic integers. 
In other words, $x$ is the expansion in an integer base $b$ of a real number between $0$ and $1$, and the odometer maps $x$ to $x + (1, 0, 0, \ldots)$ where "$+$" is the addition modulo $b$ with carry to the right. 

```{r}
odometer(c(1,0,1), base=2)
```

By transporting the $3$-adic integers to $[0,1)$ with the expansion in base $3$, let's have a look at the graph of the $3$-adic odometer:

```{r, fig.width=5, fig.height=5, fig.align='center'}
ternary2num <- function(t) sum(t/3^seq_along(t))
num2ternary <- function(u) floatExpand01(u, base=3)
par(mar=c(4,4,2,2))
u <- seq(0, 0.995, by=0.0025)
Ou <- sapply(u, function(u) ternary2num(odometer(num2ternary(u), base=3)))
plot(u, Ou, xlab="u", ylab="O(u)", xlim=c(0,1), ylim=c(0,1),
     pch=19, cex=.25, pty="s", xaxs="i", yaxs="i")
grid(nx=9)
```

The `expansions` package also provides the function `sumadic` to perform the sum of two adic integers:

```{r}
sumadic(c(0,1), c(1,1,1), base=2)
identical(sumadic(c(0,1), 1, base=2), odometer(c(0,1), base=2))
```

