---
title: "Avoiding nested loops with a Cantor expansion"
author: "Stéphane Laurent"
date: "10/08/2015"
output: html_document
---

*(first version)*


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(collapse=TRUE, fig.path="assets/fig/greedy-")
source("./assets/Rfunctions/Bgraph.R")
source("./assets/knitrEngines/V800.R")
opts_chunk$set(V8.libraries=c("assets/htmlframes/libraries/underscore-min.js"))
```


I added an answer to a [question on Stackoverflow](http://stackoverflow.com/q/4683539/1100107) about the problem of a variable amount of nested loops. But an answer to a four years old question on Stackoverflow rarely receives some attention. 

## Cantor expansions of integers

My answer is based on a Cantor expansion of the integers. 
The binary representation of integer numbers is well-known. In R, use this function to get it:

```{r number2binary}
number2binary <- function(number, noBits=1+floor(log2(max(number,1)))) {
  if(noBits < 1+floor(log2(number))) warning(sprintf("noBits=%s is not enough", noBits))
  return( as.numeric(intToBits(number))[1:noBits] ) 
}
number2binary(5)
number2binary(5, noBits=4)
```

The second argument, `noBits`, allows to fix the length of the binary representation, adding some `0` to the right to reach it.

The binary representation is a system of numerotation. The correspondence between the usual decimal representation and the binary representation is the following:

```{r}
nmax <- 10
data.frame(
  N=0:nmax,
  expansion=vapply(0:nmax, function(i) paste(myutils::number2binary(i), collapse=""), character(1))
)
```

The numerotation for the $2^n$ first integers is also given by the Cartesian product of the set $\{0,1\}$ with itself $n$ times:

```{r}
expand.grid(epsilon0=c(0,1), epsilon1=c(0,1), epsilon2=c(0,1))
```

Thus, one can view the binary representation of a number $n$ as the $n$-th element of this Cartesian product (for the [colexicographic order](https://en.wikipedia.org/wiki/Lexicographical_order#Colexicographic_order)). The enumeration is easily viewed on a tree:

```{r, echo=FALSE, fig.width=6, fig.height=3}
par(mar=c(1,1,1,1)/100)
Tree_plot(c(2,2,2), N=3, col_path="gray", lwd_path=2, only_end=TRUE, hor=FALSE)
```


Well, everybody knows that. In mathematical terms, 
 the binary representation allows to write any natural number $N$ as a finite sequence of 
digits $\epsilon_i \in \{0,1\}$:
$$
N = \text{"}\epsilon_0\epsilon_1\epsilon_2\ldots\epsilon_n\text{"}.
$$
and that means
$$
N = \epsilon_0 + \epsilon_1\times 2 + \epsilon_2 \times 2^2 + \ldots + \epsilon_n \times 2^n.
$$

Such a representation is more generally possible with digits $\epsilon_i \in \{0,\ldots, r_n-1\}$, where $(r_n)$ is a sequence of integer numbers $\geq 2$, and then the representation means
$$
N = \epsilon_0 + \epsilon_1\times \ell_1 + \epsilon_2 \times \ell_2 + \ldots + \epsilon_n \times \ell_n,
$$
where $\ell_n=\prod_{i=0}^{n-1} r_i$, and then this expansion of $N$ is called its $(r_n)$-ary expansion. The binary expansion is the case when $r_n\equiv 2$. The general $(r_n)$-ary expansion is called a Cantor's number system. There also are, beyond the scope of this post,  [more general systems of numeration](http://matwbn.icm.edu.pl/ksiazki/aa/aa70/aa7022.pdf).

For example, the $(3,4,5)$-ary representation of a number $n$ is the $n$-th element of this Cartesian product $\{0,1,2\}\times\{0,1,2,3\}\times\{0,1,2,3,4\}$.

One gets the integer $N$ from its $(r_n)$-ary expansion by simply applying the above formula. Conversely, the derivation of the $(r_n)$-ary from an integer is given by the *greedy algorithm*:

* While $N>0$
    - Take the first index $k$ such that $\ell_k > N$
    - Set $\epsilon_k$ to be the Euclidean quotient of $N$ by $\ell_{k-1}$, and update $N$ to be the remainder 

This R function returns the $(r_n)$-ary expansion, using the binary expansion as the default one:

```{r}
intToAry <- function(n, sizes=rep(2, 1+floor(log2(max(n,1))))){
  l <- c(1, cumprod(sizes))
  epsilon <- numeric(length(sizes))
  while(n>0){
    k <- which.min(l<=n)
    e <- floor(n/l[k-1])
    epsilon[k-1] <- e
    n <- n-e*l[k-1]
  }
  return(epsilon)
}
aryToInt <- function(epsilon, sizes=rep(2, length(epsilon)-1)){
  sum(epsilon*c(1,cumprod(sizes[1:(length(epsilon)-1)])))
}
```

As an example:
```{r}
intToAry(29, c(3, 4, 5)) 
```
means that $29 = 2\times 1 + 1 \times 3  + 2 \times (3\times 4)$. 

Note that the last element of `sizes`, here `5`, is not visible in the expansion. It 
only fixes the maximal value of $\epsilon_2$: the $(3,4,5)$-ary expansion of $N$ is 
$$
N = \epsilon_0 + \epsilon_1 \times 3 + \epsilon_2 \times (3\times 4)
$$
with $\epsilon_0 \in \{0,1,2\}$, $\epsilon_1 \in \{0, 1, 2, 3\}$, $\epsilon_2 \in \{0, 1, 2, 3, 4\}$.


## Application to nested loops 

Assume you have a nested loop:

```{r, eval=FALSE}
for(i in 1:3){
  for(j in 1:4){
    for(k in 1:5){
      doSomething(c(i,j,k))
    }
  }
}
```

This nested loop can be reduced to only one loop with the $(3,4,5)$-ary representation:

```{r, eval=FALSE}
for(n in 1:(3*4*5)){
  doSomething(intToAry(n)+1, c(3,4,5))
}
```


## A Rcpp implementation 

Using the `intToAry` function in a long loop could be time-consuming. We give below a `C++` implementation for R, using the `Rcpp` and `inline` packages.

```{r}
library(inline)
src <- 'int n = as<int>(N);
std::vector<int> s = as< std::vector<int> >(sizes);
std::vector<int> epsilon (s.size());
std::vector<int>::iterator it;
it = s.begin();
it = s.insert ( it , 1 );
int G[s.size()];
std::partial_sum (s.begin(), s.end(), G, std::multiplies<int>());
int k;
while(n>0){
  k=1;
  while(G[k]<=n){
    k=k+1;
  }
  epsilon[k-1] = (int)n / G[k-1];
  n = n % G[k-1];
}
return wrap(epsilon); '
intToAry_Rcpp <- cxxfunction(signature(N="integer",
                           sizes="integer"),
                 body=src, plugin="Rcpp")
```


```{r}
intToAry_Rcpp(29, c(3,4,5))
```

### Benchmarks 

The `C++` implementation is clearly faster.

```{r}
L <- vector(mode="list", length=2*3*4*5*6*7*8*9)
system.time(
  for(n in 1:(2*3*4*5*6*7*8*9)){
    L[[n]] <- intToAry(n-1, c(2,3,4,5,6,7,8,9))
  }
)
system.time(
  for(n in 1:(2*3*4*5*6*7*8*9)){
    L[[n]] <- intToAry_Rcpp(n-1, c(2,3,4,5,6,7,8,9))
  }
)
```

## A Javascript implementation 

```{r, engine='V8'}
function intToAry(n, sizes) {
  if (n<0) throw new Error("n must be a nonnegative integer");
  for (i = 0; i<sizes.length; i++) if (sizes[i]!=parseInt(sizes[i]) || sizes[i]<1){  throw new Error("sizes must be a vector of integers be >1"); };
  for (var G = [1], i = 0; i<sizes.length; i++) G[i+1] = G[i] * sizes[i]; 
  if (n>=_.last(G)) throw new Error("n must be <" + _.last(G));
  for (var epsilon=[], i=0; i < sizes.length; i++) epsilon[i]=0;
  while(n > 0){
    var k = _.findIndex(G, function(x){ return n < x; }) - 1;
    var e = (n/G[k])>>0;
    epsilon[k] = e;
    n = n-e*G[k];
  }
  return epsilon;
}
intToAry(29, [3, 4, 5])
```




