---
title: "Dataframes in Haskell: the Frames library"
date: "2016-08-24"
---

<link rel="stylesheet" href="hscolour.css">

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "./assets/haskell/")
## a small modification of the haskell engine provided by knitr:
##   (to get rid of the option ':set +m')
knitr::knit_engines$set(ghc = function (options) 
                  {
                    engine = options$engine
                    f = basename(tempfile(engine, ".", ".txt"))
                    writeLines(options$code, f)
                    on.exit(unlink(f))
                    code = paste(f, options$engine.opts)
                    cmd = options$engine.path 
                    out = if (options$eval) {
                      message("running: ", cmd, " ", code)
                      tryCatch(system2(cmd, code, stdout = TRUE, stderr = FALSE,
                                       env = options$engine.env), error = function(e) {
                                         if (!options$error)
                                           stop(e)
                                         paste("Error in running command", cmd)
                                       })
                    }
                    else ""
                    if (!options$error && !is.null(attr(out, "status"))) 
                      stop(paste(out, collapse = "\n"))
                    knitr::engine_output(options, options$code, out)
                  }
)
## chunk options
knitr::opts_chunk$set(engine = 'ghc', 
                      engine.path = 'ghcscriptrender', 
                      engine.opts = '--fragment -p .cabal-sandbox/x86_64-linux-ghc-7.10.3-packages.conf.d',
                      echo = FALSE,
                      results = 'asis')
```


This article demonstrates the use of the `Frames` library. 
It has three parts, each one containing a `ghc` script. We'll load the following module at the beginning of each script.

<div class='sourceCode'><pre class='scriptHaskell'><code class='scriptHaskell'><span class='hs-comment'>-- Frames_init.hs</span>
<span class='hs-comment'>{-# LANGUAGE TemplateHaskell #-}</span>
<span class='hs-comment'>{-# LANGUAGE DataKinds #-}</span>
<span class='hs-comment'>{-# LANGUAGE FlexibleContexts #-}</span>
<span class='hs-comment'>{-# LANGUAGE TypeOperators #-}</span>
<span class='hs-comment'>{-# LANGUAGE OverloadedStrings #-}</span>
<span class='output'> </span>
<span class='hs-keyword'>module</span> <span class='hs-conid'>Init</span> <span class='hs-keyword'>where</span> 
<span class='output'> </span>
<span class='hs-keyword'>import</span> <span class='hs-keyword'>qualified</span> <span class='hs-conid'>Control</span><span class='hs-varop'>.</span><span class='hs-conid'>Foldl</span> <span class='hs-keyword'>as</span> <span class='hs-conid'>L</span>
<span class='hs-keyword'>import</span> <span class='hs-keyword'>qualified</span> <span class='hs-conid'>Data</span><span class='hs-varop'>.</span><span class='hs-conid'>Foldable</span> <span class='hs-keyword'>as</span> <span class='hs-conid'>F</span>
<span class='hs-keyword'>import</span> <span class='hs-conid'>Data</span><span class='hs-varop'>.</span><span class='hs-conid'>Proxy</span> <span class='hs-layout'>(</span><span class='hs-conid'>Proxy</span><span class='hs-layout'>(</span><span class='hs-keyglyph'>..</span><span class='hs-layout'>)</span><span class='hs-layout'>)</span>
<span class='hs-keyword'>import</span> <span class='hs-conid'>Lens</span><span class='hs-varop'>.</span><span class='hs-conid'>Family</span> <span class='hs-layout'>(</span><span class='hs-varid'>view</span><span class='hs-layout'>,</span> <span class='hs-varid'>set</span><span class='hs-layout'>,</span> <span class='hs-varid'>over</span><span class='hs-layout'>)</span>
<span class='hs-keyword'>import</span> <span class='hs-conid'>Frames</span>
<span class='hs-keyword'>import</span> <span class='hs-conid'>Frames</span><span class='hs-varop'>.</span><span class='hs-conid'>CSV</span> <span class='hs-layout'>(</span><span class='hs-varid'>defaultParser</span><span class='hs-layout'>,</span> <span class='hs-varid'>readTableOpt</span><span class='hs-layout'>)</span> 
<span class='hs-keyword'>import</span> <span class='hs-conid'>Pipes</span> <span class='hs-varid'>hiding</span> <span class='hs-layout'>(</span><span class='hs-conid'>Proxy</span><span class='hs-layout'>)</span> <span class='hs-comment'>-- Producer</span>
<span class='hs-keyword'>import</span> <span class='hs-keyword'>qualified</span> <span class='hs-conid'>Pipes</span><span class='hs-varop'>.</span><span class='hs-conid'>Prelude</span> <span class='hs-keyword'>as</span> <span class='hs-conid'>P</span>
<span class='output'> </span>
<span class='hs-comment'>-- define a Show instance for frames</span>
<span class='hs-keyword'>instance</span> <span class='hs-layout'>(</span><span class='hs-conid'>Show</span> <span class='hs-varid'>a</span><span class='hs-layout'>)</span> <span class='hs-keyglyph'>=&gt;</span> <span class='hs-conid'>Show</span> <span class='hs-layout'>(</span><span class='hs-conid'>Frame</span> <span class='hs-varid'>a</span><span class='hs-layout'>)</span> <span class='hs-keyword'>where</span>
  <span class='hs-varid'>show</span> <span class='hs-layout'>(</span><span class='hs-conid'>Frame</span> <span class='hs-varid'>l</span> <span class='hs-varid'>f</span><span class='hs-layout'>)</span> <span class='hs-keyglyph'>=</span> <span class='hs-layout'>(</span><span class='hs-varid'>show</span> <span class='hs-varop'>$</span> <span class='hs-varid'>f</span> <span class='hs-num'>0</span><span class='hs-layout'>)</span> 
                       <span class='hs-varop'>++</span> <span class='hs-layout'>(</span><span class='hs-keyword'>if</span> <span class='hs-varid'>l</span><span class='hs-varop'>&gt;</span><span class='hs-num'>1</span> <span class='hs-keyword'>then</span> <span class='hs-str'>"\n"</span> <span class='hs-varop'>++</span> <span class='hs-layout'>(</span><span class='hs-varid'>show</span> <span class='hs-varop'>$</span> <span class='hs-varid'>f</span> <span class='hs-num'>1</span><span class='hs-layout'>)</span> <span class='hs-keyword'>else</span> <span class='hs-str'>""</span><span class='hs-layout'>)</span>
                         <span class='hs-varop'>++</span> <span class='hs-layout'>(</span><span class='hs-keyword'>if</span> <span class='hs-varid'>l</span><span class='hs-varop'>&gt;</span><span class='hs-num'>2</span> <span class='hs-keyword'>then</span> <span class='hs-str'>"\n..."</span> <span class='hs-keyword'>else</span> <span class='hs-str'>""</span><span class='hs-layout'>)</span> 
                           <span class='hs-varop'>++</span> <span class='hs-str'>"\nFrame with "</span> <span class='hs-varop'>++</span> <span class='hs-layout'>(</span><span class='hs-varid'>show</span> <span class='hs-varid'>l</span><span class='hs-layout'>)</span> <span class='hs-varop'>++</span> <span class='hs-keyword'>if</span><span class='hs-layout'>(</span><span class='hs-varid'>l</span><span class='hs-varop'>&gt;</span><span class='hs-num'>1</span><span class='hs-layout'>)</span> <span class='hs-keyword'>then</span> <span class='hs-str'>" rows."</span> <span class='hs-keyword'>else</span> <span class='hs-str'>" row."</span> 
<span class='output'> </span>
<span class='output'> </span>
<span class='hs-comment'>-- - #~ Will be used in part 1 ~# - --</span>
<span class='output'> </span>
<span class='hs-comment'>-- define new types for the records</span>
<span class='hs-definition'>declareColumn</span> <span class='hs-str'>"Name"</span> <span class='hs-chr'>'</span><span class='hs-chr'>'</span><span class='hs-conid'>String</span>
<span class='hs-definition'>declareColumn</span> <span class='hs-str'>"Value"</span> <span class='hs-chr'>'</span><span class='hs-chr'>'</span><span class='hs-conid'>Float</span>
<span class='hs-definition'>declareColumn</span> <span class='hs-str'>"Indic"</span> <span class='hs-chr'>'</span><span class='hs-chr'>'</span><span class='hs-conid'>Int</span>
<span class='hs-definition'>declareColumn</span> <span class='hs-str'>"Status"</span> <span class='hs-chr'>'</span><span class='hs-chr'>'</span><span class='hs-conid'>Bool</span>
<span class='output'> </span>
<span class='output'> </span>
<span class='hs-comment'>-- - #~ Will be used in part 2 ~# - --</span>
<span class='output'> </span>
<span class='hs-definition'>tableTypes</span> <span class='hs-str'>"Iris"</span> <span class='hs-str'>"iris_small.csv"</span>
<span class='output'> </span>
<span class='hs-definition'>irisStream</span> <span class='hs-keyglyph'>::</span> <span class='hs-conid'>Producer</span> <span class='hs-conid'>Iris</span> <span class='hs-conid'>IO</span> <span class='hs-conid'>()</span>
<span class='hs-definition'>irisStream</span> <span class='hs-keyglyph'>=</span> <span class='hs-varid'>readTableOpt</span> <span class='hs-varid'>defaultParser</span> <span class='hs-str'>"iris_small.csv"</span>
<span class='output'> </span>
<span class='hs-definition'>loadIris</span> <span class='hs-keyglyph'>::</span> <span class='hs-conid'>IO</span> <span class='hs-layout'>(</span><span class='hs-conid'>Frame</span> <span class='hs-conid'>Iris</span><span class='hs-layout'>)</span>
<span class='hs-definition'>loadIris</span> <span class='hs-keyglyph'>=</span> <span class='hs-varid'>inCoreAoS</span> <span class='hs-varid'>irisStream</span>
<span class='output'> </span>
<span class='output'> </span>
<span class='hs-comment'>-- - #~ Will be used in part 3 ~# - --</span>
<span class='output'> </span>
<span class='hs-definition'>declareColumn</span> <span class='hs-str'>"Ratio"</span> <span class='hs-chr'>'</span><span class='hs-chr'>'</span><span class='hs-conid'>Double</span>
<span class='hs-definition'>declareColumn</span> <span class='hs-str'>"MeanWidth"</span> <span class='hs-chr'>'</span><span class='hs-chr'>'</span><span class='hs-conid'>Double</span>
</code></pre></div>

Note that we have defined a `Show` instance for the dataframes, because there is none. 


## Records

The first script below demonstrates how to define records and a dataframe by hand. 
In order to define these records, we used `declareColumn` in the `Init` module. It creates new types and lenses.

```{r script1, cache=TRUE}
:load Frames_init.hs

-- - #~ Records ~# - --

-- define records
:set -XDataKinds
:{
let row1 :: Record '[Name, Value, Indic, Status]
    row1 =  "Joe" &: 10.1 &: 0 &: True &: Nil
:}
:{
let row2 :: Record '[Name, Value, Indic, Status]
    row2 =  "Bill" &: 9.0 &: 1 &: True &: Nil
:}
:{
let row3 :: Record '[Name, Value, Indic, Status]
    row3 =  "Kate" &: -3.0 &: 2 &: False &: Nil
:}

-- take a look
row1

-- declareColumn has defined new lenses name, value, indic and status
:info name

-- let's use them
view status row1 
-- it's the same as
rget status row1
-- setter: 
set name "Jim" row1
rput name "Jim" row1
-- apply a function:
over value (+1) row1 


-- - #~ Now let's make a frame ~# - --
let fr = toFrame [row1, row2, row3]
:t fr 

-- have a look thanks to our Show instance
fr

-- how many rows?
frameLength fr

-- take row 0
frameRow fr 0 == row1

-- we can convert the frame to a list
:t F.toList fr 

-- look at the Name column
view name <$> fr
-- this is a frame 
:t view name <$> fr 
-- the same as
fmap (rget name) fr
-- as well as 
rget name <$> fr

-- set every Status to False 
rput status False <$> fr

-- double every Value
over value (*2) <$> fr

-- get the minimal Value
L.fold L.minimum (rget value <$> fr)
-- or 
minimum $ F.toList (rget value <$> fr)
```

Note that our imports of the functions `view` and `set` of the `Lens.Family` module were not necessary because we can use `rget` and `rput` instead (imported from the `Frames.RecLens` module).


## CSV import 

The `Frames` library allows to get a dataframe from a CSV file. The import (in the `Init` module) generates new types of records for the rows, and lenses as before.

The CSV file we use contains the following small subset of the well-know iris dataset. 

```{r iris, engine='R', results='markup'}
iris <- read.csv("iris_small.csv")
knitr::kable(iris, row.names=FALSE, align="c")
```

```{r script2, cache=TRUE}
:load Frames_init.hs 

-- - #~  tableTypes results ~# - --

-- new types: Id, PetalLength, PetalWidth, and Species
:info PetalLength

-- lenses: id, petalLength, petalWidth, and species
:t species 

-- Iris: an alias for the Record type of the rows
:info Iris  


-- - #~ the iris frame ~# - --

iris <- loadIris 
iris
:t iris 
```


## Subset, mutate, filter, aggregate

Here we give some examples of common operations on dataframes. Note that the types `Ratio` and `MeanWidth` were defined in the `Init` module.

```{r script3, cache=TRUE}
:load Frames_init.hs 
iris <- loadIris 


-- - #~ subset of rows ~# - --

toFrame $ map (frameRow iris) [3, 4, 5] 

-- - #~ subset of columns ~# - --
-- select Id and Species of row 3
:set -XQuasiQuotes
:set -XDataKinds
select [pr|Id,Species|] $ frameRow iris 3
-- select Id and Species of the frame
select [pr|Id,Species|] <$> iris


-- - #~ add a new column ~# - --

-- say we want the ratio PetalLength/PetalWidth in a new column
-- first define a function for one record
:set -XDataKinds
:{
let ratio :: Iris -> Record '[Id, PetalLength, PetalWidth, Species, Ratio]
    ratio x = rappend x ( (rget petalLength x)/(rget petalWidth x) &: Nil )
:}
-- test the ratio function on a row
ratio $ frameRow iris 0
-- map it to the frame 
fmap ratio iris


-- - #~ filtering ~# - --

-- Species is Text so we set OverloadedStrings
:set -XOverloadedStrings 
-- function returning True if 'kind' matches the species of 'record'
:{
let isThisSpecies :: Text -> Iris -> Bool
    isThisSpecies kind record = (rget species record) == kind
:}
-- test our function on a row
isThisSpecies "virginica" $ frameRow iris 0 
-- filter
filterFrame (isThisSpecies "virginica") iris


-- - #~ aggregation ~# - --

-- say we want the mean of PetalWidth for each species
-- first write a function aggregating for a given species
import Numeric.Statistics (mean)
:{
let recordMeanWidthBySpecies :: Frame Iris -> Text -> Record '[Species, MeanWidth]
    recordMeanWidthBySpecies frame kind = kind &: average &: Nil
      where average = mean . F.toList 
                        $ (rget petalWidth <$> filterFrame (isThisSpecies kind) frame) 
:}
-- test on one species
recordMeanWidthBySpecies iris "versicolor" 
-- now map this function to all species, and frame
import Data.List (nub)
let allspecies = nub . F.toList $ rget species <$> iris
toFrame $ map (recordMeanWidthBySpecies iris) allspecies 

-- put things together 
:{
let meanWidthAggregationBySpecies :: Frame Iris -> Frame (Record '[Species, MeanWidth])
    meanWidthAggregationBySpecies frame = toFrame $ map (recordMeanWidthBySpecies frame) allspecies
      where allspecies = nub . F.toList $ rget species <$> frame
            recordMeanWidthBySpecies fr kind = kind &: average &: Nil
              where average = mean . F.toList 
                                $ (rget petalWidth <$> filterFrame (isThisSpecies kind) fr)
:}
meanWidthAggregationBySpecies iris
```

