---
title: "Ruby is cool"
author: "Stéphane Laurent"
date: "22/06/2015"
output: html_document
---

*(first version)*

```{r, include=FALSE, memo=TRUE}
library(knitr)
join <- function(x){ 
  tab <- stringr::str_detect(x, "  ") | stringr::str_detect(x, "\t")
  begins <- which(sapply(seq_len(length(x)-1), function(i) !tab[i] && tab[i+1])) 
  ends <- which(sapply(seq_len(length(x)-1), function(i) tab[i] && !tab[i+1]))+1
  for(i in 1:length(begins)){
    begin <- begins[i]
    end <- ends[i]
    x[begin] <- paste(x[begin:end], collapse="\n")
    x[(begin+1):end] <- NA
    }
  x <- x[-which(is.na(x))]
  return(x)
  }
# example: 
# > join(c("x=2", "y=3;", "def f(a)", "  w=a*a",  "  return w", "end", "puts f(x+y)", "def g(a)", "  return a*a*a", "end"))
# [1] "x=2"                                "y=3;"                              
# [3] "def f(a)\n  w=a*a\n  return w\nend" "puts f(x+y)"                       
# [5] "def g(a)\n  return a*a*a\nend"   
knit_engines$set(RUBY = function(options) {
  if(!exists("MEMO")) MEMO <<- list()
  if(length(MEMO) && !is.null(options$use.memo) && options$use.memo){
    precode <- do.call(c, MEMO)
    memo <- TRUE
    }else{
      memo <- FALSE
      }
  code0 <- options$code
  if(!is.null(options$memo) && options$memo) MEMO <<- c(MEMO, list(code0))
  if(length(code0)>1){
    code <-   if(any(stringr::str_detect(code0, "  "))) join(code0) else code0
    }else{
      code <- code0
      }
  prog <- paste(sapply(code, function(x) sprintf("puts '>> %s'\n%s", x, x)), collapse="\n")
  if(memo){
    prog <- paste(paste(precode, collapse="\n"), "\n", prog, collapse="\n")
    }
  out <-  system2("ruby", args=c("-e", shQuote(prog)), stdout=TRUE, stderr=TRUE)
  start <- which.max(out==paste0(">> ", code0[1]))
  if(memo) out <- out[start:length(out)] # tail(out, length(out)-start+2)
  out <- paste(out, collapse="\n") #  capture.output(str(out))
  options$comment=NA
  return( knitr:::wrap.character(out, options))
  })
```


Do you think this title is ridiculous ? I do. But last week I gave my first try to [Ruby](https://www.ruby-lang.org/en/), and I have nothing really useful to share: this blog post is just the story of someone discovering Ruby.

I was looking for a (cool) programming language providing a way to deal with sets. The [set library](http://ruby-doc.org/stdlib-2.2.2/libdoc/set/rdoc/Set.html) perfectly fulfilled my expectation :

```{r, engine='RUBY', memo=TRUE}
require "set" 
s1 = ["a", "b"].to_set
puts s1.inspect
s2 = ["b", "c"].to_set
puts s2.inspect
puts s1.subset? s2
puts s1.union(s2).inspect
```

Is there something to explain in the above code ? I don't think so (unless you have never used any programming language). And this one of the reasons for which Ruby is cool.

## Set-theoretic construction of natural numbers

In order to train myself to use the Ruby langage, I undertook to implement the set-theoretic construction of the natural numbers:
$$0 = \{\varnothing\}, \quad 1=0 \cup \{0\}= \{\varnothing,  \{\varnothing\}\}, \quad 2=1\cup\{1\}=\{\varnothing,  \{\varnothing\}, \{\varnothing,  \{\varnothing\}\}\}, \quad ...$$

To define $0$ we can use $\varnothing$ as a character or as a symbol:

```{r, engine='RUBY', use.memo=TRUE}
puts ["Ø"].to_set.inspect
puts [:Ø].to_set.inspect
```

Now, we define a recursive function calculating the set representation of an integer:

```{r, engine='RUBY', use.memo=TRUE, memo=TRUE}
def set_representation(n) 
  n == 0 ? ["Ø"].to_set : set_representation(n-1).tap {|x| break x.union([x].to_set)} 
end
```

See [this stackoverflow thread](http://stackoverflow.com/questions/14780783/ruby-map-for-a-single-object) about the usage of `tap` to map a single object to a function. 

The `set_representation`  function works fine:

```{r, engine='RUBY', use.memo=TRUE}
puts set_representation(1).inspect
puts set_representation(2).inspect
```

However, another way consists in using a *hash*, as I learnt by reading [Algorithmic Fun with Ruby Hashes](http://www.sitepoint.com/algorithmic-fun-ruby-hashes/). The code is similar to the one of `set_representation`, except that the first value must be initialized outside the body:

```{r, engine='RUBY', memo=TRUE, use.memo=TRUE}
set_hash ||= Hash.new do |hash,n|
  hash[n] =  hash[n-1].tap {|x| break x.union([x].to_set)}
end
# initialization:
set_hash[0] = ["Ø"].to_set
```

Results, as expected, are the same:

```{r, engine='RUBY', use.memo=TRUE}
puts set_hash[2].inspect
```

And the computation times are comparable:

```{r, engine='RUBY', use.memo=TRUE, memo=TRUE}
# time consumed by the function:
start = Time.now; set_representation(15); finish = Time.now
puts finish - start
# time consumed by the hash:
start = Time.now; set_hash[15]; finish = Time.now
puts finish - start
```

***But*** look what happens if I re-run the same code:

```{r, engine='RUBY', use.memo=TRUE, memo=FALSE}
# time consumed by the function:
start = Time.now; set_representation(15); finish = Time.now
puts finish - start
# time consumed by the hash:
start = Time.now; set_hash[15]; finish = Time.now
puts finish - start
```

Now, the time consumed by `set_hash[15]` is almost instantaneous! This is due to [*memoization*](https://en.wikipedia.org/wiki/Memoization) (see  [Algorithmic Fun with Ruby Hashes](http://www.sitepoint.com/algorithmic-fun-ruby-hashes/)). 


## Displaying a set 

My next goal was to get a better rendering of the sets. Indeed, it is hard to read something like :

```{r, engine='RUBY', use.memo=TRUE}
puts set_hash[2].inspect
```

as compared to $$\{\varnothing,  \{\varnothing\}, \{\varnothing,  \{\varnothing\}\}\}$$ isn't it ? 
Then I defined a recursive hash:

```{r, engine='RUBY', use.memo=TRUE, memo=TRUE}
display_hash ||= Hash.new do |hash,set|
  if set.inject(true) {|status, x| status and x.class==String}
    display_hash[set] = "{" + set.inject {|union, x| union + "," + x} + "}"
  else 
    set = set.to_a
    for i in 0 .. set.size-1
      if set[i].class == Set 
        set[i] = display_hash[Set.new(set[i])]
      end
    end
    display_hash[set] = display_hash[set.to_set] 
  end
end
```

The first line checks whether all elements of the set are some strings. If so, it returns the set as a character string with a comma separating the elements and some braces. Otherwise, the hash is calculated for each element of the set which are themselves some sets (the procedure assumes that the set to be displayed consists only of character strings and of sets of such sets). 
And it works as desired:

```{r, engine='RUBY', use.memo=TRUE}
one = set_hash[1]
two = set_hash[2]
three = set_hash[3]
puts one.inspect # ugly 
puts display_hash[one] # beautiful
puts two.inspect # ugly
puts display_hash[two] # beautiful
puts three.inspect # ugly 
puts display_hash[three] # beautiful
```

The memoization offered by hash approach has a great advantage here. Take for example the case of $2$:
$$2 = \{\varnothing,  \{\varnothing\}, \{\varnothing,  \{\varnothing\}\}\} = \{\varnothing,  \{\varnothing\}, 1\},$$
and then, when the set representation of $2$ is generated, the set representation of $1$ is generated as well, and this one is memoized. 
For example we can see that the time consumed by `set_hash[16]` is almost instantaneous once  `set_hash[17]` is calculated: 

```{r, engine='RUBY', use.memo=TRUE, memo=FALSE}
# time consumed by set_hash[17]:
start = Time.now; set_hash[17]; finish = Time.now
puts finish - start
# time consumed by set_hash[16]:
start = Time.now; set_hash[16]; finish = Time.now
puts finish - start
```

## Unordered pairs 

Finally, I wanted to visualize the iterated unordered pairs of a set. That is, starting for example from the set $A_1=\{a,b\}$, the set of its unordered pairs is 
$A_2 = \{\{a\}, \{a,b\}, \{b\}\}$, then we form the set of unordered pairs of $A_2$, and so on. A simple function does this job:

```{r, engine='RUBY', use.memo=TRUE, memo=TRUE}
def pairs(s,n)
  while n > 0 do 
		b = s.to_a
		s = [].to_set
		for i in 0 .. b.size-1
			for j in i .. b.size-1
				s.add([b[i],b[j]].to_set)
			end
		end
		n = n - 1
	end
	return s.to_set
end
```

And the rendering hash works fine: 

```{r, engine='RUBY', use.memo=TRUE}
p1 = pairs(["a","b"].to_set, 1)
puts display_hash[p1]
p2 = pairs(["a","b"].to_set, 2)
puts display_hash[p2]
```

We can also list the iterated pairs using `map`:

```{r, engine='RUBY', use.memo=TRUE}
p3 = pairs(["a","b"].to_set, 3)
puts p3.map {|x| display_hash[x]}
```

<br/> 

Now, maybe you still think the title is ridiculous. But I hope you believe Ruby is cool.


```{r, echo=FALSE}
rm(list=ls())
knit_exit()
```
